Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex
Resectioned a little & more edits.

Fortunately, the genotype matrix is not the only way to encode genetic data.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
structure that efficiently encodes the genealogical trees describing how a set of
individuals are related to each other at each point along their genome.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, and led to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similar efficiency gains.
Recent work~\citep{kelleher2019inferring} has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples.
The key to the remarkable scalability and efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, while also
allowing us to to compute several population genetics statistics
efficiently~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics.

@@

% summary
In this paper, we study only ``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site,
leaving other statistics (e.g., pairwise statistics such as linkage disequilibrium
or haplotype statistics) for later work.
The methods we describe are implemented, tested, and validated in the \tskit{} library,
which is available as an open-source Python package at \url{https://github.com/tskit-dev/tskit}.

@@

We will use weights that count number of samples inheriting from each node
sufficiently often that we give them a special name --
for a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

Now we describe how to compute statistics from \emph{genomes} using this framework.
To do this, we assume that the genetic variation data is embedded in a tree sequence,
but the trees are used only for efficiency --
the results will not depend on the trees in any way
(with the only exception being if the ancestral allele is assumed to be known, as discussed below).

The summaries we defined above for each node in each tree
extend nicely to genetic variants
-- just as a node weight contains information about which samples inherit from the node and which are not,
so we can summarise patterns of genetic variation by summing up weights of all samples
that carry a given allele.
So, we define \emph{allele weights} to be
the total weight of all sample nodes who have inherited that allele.

@@

If there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
More generally, $\aw$ is not necessarily equal to a subtree weight
but is easily computable from them.

Using allele weights,
we define the \textbf{site statistic} of site $j$ computed using a summary function $f()$
and a set of sample weights $\iw$
to be the sum of $f()$ applied to each of the allele weights:
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a} f(\aw_j(a)) ,
\end{align}
where the sum is over all unique alleles at site $j$.

Of course, we often want to summarize statistics across regions of the genome (``windows'').
To do this, we overload notation somewhat and use a subscript $[i,j)$ to denote an average
over the corresponding portion of the genome:
\begin{align}
    \site(f, \iw)_{[i,j)}
    &=
    \frac{1}{j-i} \sum_{k=i}^{j-1} \site(f, \iw)_k
\end{align}
Since we have required $f(0) = f(\tiw) = 0$,
the sum will only be affected by polymorphic sites,
although the normalization is always by total number of sites, to allow comparison between regions.

In the definition above we simply sum over all alleles at each site.
However, sometimes it is useful to distinguish the \emph{ancestral} allele
(i.e., the allele at the root of the tree) from the remaining derived alleles.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily),
and one way to make use of this information is to sum over only derived alleles.
Analogously to the above,
we say a site statistic is \textbf{polarised} if we do this,
defining
\begin{align} \label{eqn:site_polarised}
    \text{(polarised)} \qquad
    \sitep(f, \iw)_j
    &=
    \sum_{a \in D_j} f(\aw_j(a)) ,
\end{align}
where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
Note that this may not be quite what is expected,
for instance, if there has been a back mutation to the ancestral allele at some point in the tree,
or if there have been mutations to distinct alleles on different parts of the tree
such that the ancestral allele is no longer present.
However, since these situations depend on multiple mutations occurring at a single site,
they are relatively rare in practice.

Having defined a general framework, we may now give concrete examples
of how to compute common summary statistics

@@

@@

\begin{example}[Phenotypic correlations] \label{ex:site_correlations}
    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z(u)$,
    and we want to compute the correlation between this phenotype
    and the genotype at each site in the tree sequence.
    For convenience, suppose $z$ is normalized
    % so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2/(n-1) = 1$.
    to have mean zero and variance 1.
    Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
    then the covariance of $z$ with $g_j$ is just $\sum_u z(u) g_j(u) = \sum_{u \st g_j(u) = 1} z(u)$,
    i.e., the sum of the phenotypes of all samples carrying the derived allele,
    divided by $(n - 1)$, where $n$ is the number of samples.
    Since the phenotypes sum to zero, this is also equal to
    $- \sum_{u \st g_j(u) = 0} z(u)$.
    If $p_j = \sum_u g_j(u)$ is the derived allele frequency at site $j$,
    then the variance of $g_j$ is $n p_j (1-p_j) / (n-1)$,
    and so the squared correlation can be calculated as a sum across the two alleles:
    \begin{align*}
        r_j^2 =
        \frac{\left( \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)n(n-1)}
        + \frac{\left( \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)n(n-1)}  .
    \end{align*}
    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1/n$,
    and $f(x_1, x_2) = x_1^2 / (2 x_2 (1 - x_2) n (n-1))$:
    then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
\end{example}

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position --
i.e., the \emph{infinite sites model} --
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree of the tree sequence,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry ---
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch in each tree
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.

An example is shown in Figure~\ref{fig:branch_site_diagram},
showing how the $f_4$ site statistic assigns a weight to each mutation
depending on its' frequency in each of the four sample sets,
and how the corresponding branch statistic assigns the same weight to those branches.

@@

This defines a statistic for a single window,
but in practice it is more useful to average branch statistics
over a region of the genome,
which we do by averaging the tree statistics over the region
with probabilities proportional to the trees' spans.
We write this with similar notation:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).

The \emph{polarised} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarised)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarised statistics is less tight
than for the unpolarised ones.

@@

\begin{example}[Patterson's $f_4$] \label{ex:branch_f4}
    Suppose that the four subsets each consist of only a single sample.
    The summary function $f(x_1, x_2, x_3, x_4)$ for the $f_4$ statistic
    then assigns weight 1 to any branch that separates $x_1$ and $x_3$ from $x_2$ and $x_4$,
    and weight -1 to any branch that separates $x_1$ and $x_4$ from $x_2$ and $x_3$.
    The statistic $\branch(f, \iw)$ therefore
    gives the difference in average lengths of these two types of branch,
    averaged over choices of samples from the sample sets and averaged across the genome.
\end{example}

@@

\plr{not sure if we need this section}

@@

This relationship is true more generally.
In fact, for any region of the genome between $a$ and $b$,
with $\treeseq_{[a,b)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[a,b)}
    =
    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
\end{align}
where the expected value averages over infinite-sites mutations with rate 1
(and so lengths are measured in units of expected numbers of mutations)
but leaves the genealogies fixed.
In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
which is itself a summary of local tree shapes.
Let's unpack the assumptions here: what exactly is the mutational model?
First, we are taking the mutation rate to be 1, i.e.,
the expected number of mutations that occur on a region of the genome of length $\ell$
over $t$ units of time is equal to $\ell t$.
This just amounts to a change of units --
for instance, if genome length is measured in nucleotides,
and the probability of mutation is $\mu$ per nucleotide and per generation,
then we are measuring times in units of $1/\mu$ generations.
Second, we are assuming that the probability of per-site mutation is low enough
that no two mutations occur at the same site
-- the fact that they do, occasionally, means that this is an approximation.
Third, we are assuming that mutation rates are constant through time and across the genome.
Of course, the statement remains true if we can measure distance along the genome and time
in a way that mutation rates are constant, but how these vary is generally unknown.

Note that the expected product of two site statistics,
$\E[\site(f, \iw) \site(g, \iw')]$
is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
because they are not independent.
However, it is always possible to define a branch statistic that
gives the expected value of the product.
How to do this is described in \citet{ralph2019empirical}.

In this view,
site statistics are noisy approximations to the corresponding branch statistic
-- but, how noisy?
How big is the contribution of mutation to the overall sampling variance of a statistic?
Concretely the law of total variance lets us partition the variance of a site statistic
into the contributions of mutation and demographic noise:
\begin{align*}
    \var[\site(f, \iw)_{[a,b)}]
    &=
        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right]
        + \var\left[ \E\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right] \\
    &=
        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right]
        + \var\left[ \branch(f, \iw)_{[a,b)} \right] .
\end{align*}
The first term, the variance of the statistic given the trees,
is the contribution to variance of mutation,
while the second term, the variance of the branch statistic,
is the contribution of demographic variation, including drift.
In the next section we examine this partitioning of variance using simulations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Branch statistics for population genetics}

We examined the contributions of mutation and demography to noise
in two simulations where population genetic (site) statistics
are important in making inferences:
detecting recent selective sweeps,
and detecting introgression.

Figure~\ref{fig:sweep_duality} shows windowed
diversity along the chromosome following a few selective sweeps.
The top two plots compare ``branch'' diversity -- i.e., as computed only with tree shape --
to ``site'' diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
and approaches ``branch'' diversity.
These were computed using the entire population of 1,000 individuals;
how does sampling variance contribute?
Not much, it turns out -- the bottom plot shows both site and branch diversity
computed from 20 nonoverlapping groups of 100 samples.
Neither site or branch diversity vary much between these samples,
implying that the subsample gives us a good estimate of the whole-population values of each.
However, as we see in the top figure,
whole-population site diversity is itself only a quite noisy estimator of branch diversity.
The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
for a simulation with 10,000 individuals, which shows similar patterns.
Only the last of these plots is possible to directly observe in real data:
in the top two plots,
the spread of independent replicates of mutational noise (black lines)
about their expectation based on the tree sequence (red line)
is unobservable, although estimable.

\begin{figure}
    \includegraphics{{figures/swept.1000.999.1e-09.diversity}.pdf}
    \caption{
        Mean genetic diversity and time to most recent common ancestor
        in 500Kb windows along a 50Mb genome (0.5 Morgans) following several selective sweeps.
        In each case, ``site'' is mean genetic diversity (Tajima's $\pi$) divided by mutation rate,
        and ``branch'' is the corresponding branch statistic described above.
        The tree sequence was produced by simulating mutations under positive selection with mutation rate $10^{-12}$
        in a population of size 1000 for 400 generations using SLiM,
        followed by recapitation with $N_e=1000$.
        The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
        and are at frequencies 96.8\%, 100\%, 16.25\%, 100\%, and 82.6\% in the final generation, respectively.
        All curves use the same tree sequence, including selected mutations,
        and with additional neutral mutations added.
        \textbf{(top)}
        Diversity within the entire population,
        computed using for mode ``site'' from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
        \textbf{(middle)}
        As in the top panel, but with mutation rate $\mu=10^{-8}$,
        showing that as mutation rate increases, mode ``site'' (divided by $\mu$) converges to mode ``branch''.
        \textbf{(bottom)}
        ``Site'' and ``branch'' diversity within 20 disjoint samples of size 100 each,
        computed on a single tree sequence with mutation rate $\mu = 10^{-9}$.
        \label{fig:sweep_duality}
    }
\end{figure}

@@

\begin{figure}
    \includegraphics{{figures/intro/introgressed.1000.23.10000.1e-09.f4sd}.pdf}
    \caption{
        Standard deviation of site (black) and branch (red) $f_4(1,2;3,4)$
        in windows along a 100Mb genome, as a function of
        \textbf{(left)} mutation rate,
        \textbf{(middle)} sample size, and
        \textbf{(right)} window width.
        The recombination rate $10^{-8}$,
        and unless otherwise stated, the mutation rate was $10^{-9}$,
        the window width was 10Kb,
        and the entire population (1000 diploids in each of four populations) were used.
        \label{fig:introf4}
    }
\end{figure}

>>>>

Fortunately, the genotype matrix is not the only way to encode genetic data.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
structure that efficiently encodes the genealogical trees describing how a set of
individuals are related to each other at each point along their genome.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, and led to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similar efficiency gains.
Recent work~\citep{kelleher2019inferring} has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples.
The key to the remarkable scalability and efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, while also
allowing us to to compute several population genetics statistics
efficiently~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

% summary
In this paper, we study only ``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site,
leaving other statistics (e.g., pairwise statistics such as linkage disequilibrium
or haplotype statistics) for later work.
The methods we describe are implemented, tested, and validated in the \tskit{}
Python and C libraries, available at \url{https://github.com/tskit-dev/tskit}
under the terms of the MIT open source license.

@@

We will use weights that count number of samples inheriting from each node
sufficiently often that we give them a special name ---
for a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

Now we describe how to compute statistics from \emph{genomes} using this framework.
To do this, we assume that the genetic variation data is embedded in a tree sequence,
but the trees are used only for efficiency --- the results will not depend on
the trees in any way. The summaries we defined above nodes
extend directly to genetic variants: just as a node weight contains information about
which samples inherit from the node and which do not,
so we can summarise patterns of genetic variation by summing up weights of all samples
that carry a given allele.
Therefore, we define \emph{allele weights} to be
the total weight of all sample nodes that have inherited that allele.

@@

If there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
Similarly, in the more general case of recurrent and back mutations, $\aw$ can
be computed from subtree weights in a straightforward way.
We then define the \textbf{site statistic} of site $j$ for a summary function $f()$
and sample weights $\iw$ to be the sum of $f()$ applied to the weight of
each allele, $a$:
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a} f(\aw_j(a)).
\end{align}
We often want to summarize statistics across regions of the genome (``windows'').
To do this, we overload notation somewhat and use a subscript $[i,j)$ to denote an average
over the corresponding portion of the genome:
\begin{align}
    \site(f, \iw)_{[i,j)}
    &=
    \frac{1}{j-i} \sum_{k=i}^{j-1} \site(f, \iw)_k
\end{align}
Since we have required $f(0) = f(\tiw) = 0$,
the sum will only be affected by polymorphic sites,
although the normalization is always by total number of sites, to allow comparison between regions.

In the definition above we sum over all alleles at each site.
However, sometimes it is useful to distinguish the \emph{ancestral} allele
(i.e., the allele at the root of the tree) from the remaining derived alleles.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily),
and one way to make use of this information is to sum over only derived alleles.
Analogously to the above,
we say a site statistic is \textbf{polarised} if we do this,
defining
\begin{align} \label{eqn:site_polarised}
    \text{(polarised)} \qquad
    \sitep(f, \iw)_j
    &=
    \sum_{a \in D_j} f(\aw_j(a)) ,
\end{align}
where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
Note that this may not be quite what is expected,
for instance, if there has been a back mutation to the ancestral allele at some point in the tree,
or if there have been mutations to distinct alleles on different parts of the tree
such that the ancestral allele is no longer present.
However, since these situations depend on multiple mutations occurring at a single site,
they are relatively rare in practice.

Having defined a general framework, we may now give concrete examples
of how to compute common summary statistics.

@@

\jk{I don't think there's anything to be gained by giving this example over
an interval rather than saying ``Then $\site(f, \iw)_{j}$ is the nucleotide
diversity within $S$ at site $j$". The windowing stuff is routine, I think, and is just
distracting from the interesting stuff here. Ditto the next two examples.}

@@

\begin{example}[Phenotypic correlations] \label{ex:site_correlations}
    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z(u)$,
    and we want to compute the correlation between this phenotype
    and the genotype at each site.
    For convenience, suppose $z$ is normalized
    % so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2/(n-1) = 1$.
    to have mean zero and variance 1.
    Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
    then the covariance of $z$ with $g_j$ is just $\sum_u z(u) g_j(u) = \sum_{u \st g_j(u) = 1} z(u)$,
    i.e., the sum of the phenotypes of all samples carrying the derived allele,
    divided by $(n - 1)$, where $n$ is the number of samples.
    Since the phenotypes sum to zero, this is also equal to
    $- \sum_{u \st g_j(u) = 0} z(u)$.
    If $p_j = \sum_u g_j(u)$ is the derived allele frequency at site $j$,
    then the variance of $g_j$ is $n p_j (1-p_j) / (n-1)$,
    and so the squared correlation can be calculated as a sum across the two alleles:
    \begin{align*}
        r_j^2 =
        \frac{\left( \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)n(n-1)}
        + \frac{\left( \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)n(n-1)}  .
    \end{align*}
    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1/n$,
    and $f(x_1, x_2) = x_1^2 / (2 x_2 (1 - x_2) n (n-1))$:
    then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
\end{example}

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position ---
i.e., the \emph{infinite sites model} ---
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry ---
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch in each tree
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}\label{eqn:branch_stat_tree}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.
An example of this correspondence between site and branch statistics
is shown in Figure~\ref{fig:branch_site_diagram}. Here, we
see how the $f_4$ site statistic assigns a weight to each mutation
depending on its frequency in each of the four sample sets,
and how the corresponding branch statistic assigns the same weight to those branches.

@@

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarised} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarised)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarised statistics is less tight
than for the unpolarised ones.

@@

\begin{example}[Patterson's $f_4$] \label{ex:branch_f4}
    Suppose that the four subsets each consist of only a single sample.
    The summary function $f(x_1, x_2, x_3, x_4)$ for the $f_4$ statistic
    then assigns weight 1 to any branch that separates $x_1$ and $x_3$ from $x_2$ and $x_4$,
    and weight -1 to any branch that separates $x_1$ and $x_4$ from $x_2$ and $x_3$.
    The statistic $\branch(f, \iw)$ therefore
    gives the difference in average lengths of these two types of branch,
    averaged over choices of samples from the sample sets and averaged across the genome.
    \jk{I don't follow this last bit: how can we averate over choices in the
sample sets if there's only one element in each?}
\end{example}

@@

\plr{not sure if we need this section}
\jk{Compress and move to the discussion?}

@@

This relationship is true more generally.
In fact, for any region of the genome between $a$ and $b$,
with $\treeseq_{[a,b)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[a,b)}
    =
    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
\end{align}
where the expected value averages over infinite-sites mutations with rate 1
(and so lengths are measured in units of expected numbers of mutations)
but leaves the genealogies fixed.
In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
which is itself a summary of local tree shapes.
Let's unpack the assumptions here: what exactly is the mutational model?
First, we are taking the mutation rate to be 1, i.e.,
the expected number of mutations that occur on a region of the genome of length $\ell$
over $t$ units of time is equal to $\ell t$.
This just amounts to a change of units ---
for instance, if genome length is measured in nucleotides,
and the probability of mutation is $\mu$ per nucleotide and per generation,
then we are measuring times in units of $1/\mu$ generations.
Second, we are assuming that the probability of per-site mutation is low enough
that no two mutations occur at the same site
--- the fact that they do, occasionally, means that this is an approximation.
Third, we are assuming that mutation rates are constant through time and across the genome.
Of course, the statement remains true if we can measure distance along the genome and time
in a way that mutation rates are constant, but how these vary is generally unknown.

Note that the expected product of two site statistics,
$\E[\site(f, \iw) \site(g, \iw')]$
is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
because they are not independent.
However, it is always possible to define a branch statistic that
gives the expected value of the product~\citep{ralph2019empirical}.

In this view,
site statistics are noisy approximations to the corresponding branch statistic
--- but, how noisy?
How big is the contribution of mutation to the overall sampling variance of a statistic?
The law of total variance lets us partition the variance of a site statistic
into the contributions of mutation and demographic noise:
\begin{align*}
    \var[\site(f, \iw)_{[a,b)}]
    &=
        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right]
        + \var\left[ \E\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right] \\
    &=
        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right]
        + \var\left[ \branch(f, \iw)_{[a,b)} \right] .
\end{align*}
The first term, the variance of the statistic given the trees,
is the contribution to variance of mutation,
while the second term, the variance of the branch statistic,
is the contribution of demographic variation, including drift.

% JK: deleted the section here as it seemed superfluous.

We examined the contributions of mutation and demography to noise
in two simulations where population genetic (site) statistics
are important in making inferences:
detecting recent selective sweeps,
and detecting introgression.
Figure~\ref{fig:sweep_duality} shows windowed
diversity along the chromosome following a few selective sweeps.
The top two plots compare ``branch'' diversity --- i.e., as computed only with tree shape ---
to ``site'' diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
and approaches ``branch'' diversity.
These were computed using the entire population of 1,000 individuals;
how does sampling variance contribute?
Not much, it turns out --- the bottom plot shows both site and branch diversity
computed from 20 nonoverlapping groups of 100 samples.
Neither site or branch diversity vary much between these samples,
implying that the subsample gives us a good estimate of the whole-population values of each.
However, as we see in the top figure,
whole-population site diversity is itself only a quite noisy estimator of branch diversity.
The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
for a simulation with 10,000 individuals, which shows similar patterns.
Only the last of these plots is possible to directly observe in real data:
in the top two plots,
the spread of independent replicates of mutational noise (black lines)
about their expectation based on the tree sequence (red line)
is unobservable, although estimable.

\begin{figure}
    \includegraphics{{figures/swept.1000.999.1e-09.diversity}.pdf}
    \caption{
        Mean genetic diversity and time to most recent common ancestor
        in 500Kb windows along a 50Mb genome (0.5 Morgans) following several selective sweeps.
        In each case, ``site'' is mean genetic diversity (Tajima's $\pi$) divided by mutation rate,
        and ``branch'' is the corresponding branch statistic described above.
        The tree sequence was produced by simulating mutations under positive selection with mutation rate $10^{-12}$
        in a population of size 1000 for 400 generations using SLiM,
        followed by recapitation with $N_e=1000$~\citep{haller2018tree}.
        The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
        and are at frequencies 96.8\%, 100\%, 16.25\%, 100\%, and 82.6\% in the final generation, respectively.
        All curves use the same tree sequence, including selected mutations,
        and with additional neutral mutations added.
        \textbf{(top)}
        Diversity within the entire population,
        computed using for mode ``site'' from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
        \textbf{(middle)}
        As in the top panel, but with mutation rate $\mu=10^{-8}$,
        showing that as mutation rate increases, mode ``site'' (divided by $\mu$) converges to mode ``branch''.
        \textbf{(bottom)}
        ``Site'' and ``branch'' diversity within 20 disjoint samples of size 100 each,
        computed on a single tree sequence with mutation rate $\mu = 10^{-9}$.
        \label{fig:sweep_duality}
    }
\end{figure}

@@

\begin{figure}
    \includegraphics{{figures/intro/introgressed.1000.23.10000.1e-09.f4sd}.pdf}
    \caption{
        Standard deviation of site (black) and branch (red) $f_4(1,2;3,4)$
        in windows along a 100Mb genome, as a function of
        \textbf{(left)} mutation rate,
        \textbf{(middle)} sample size, and
        \textbf{(right)} window width.
        The recombination rate was $10^{-8}$,
        and unless otherwise stated, the mutation rate was $10^{-9}$,
        the window width was 10Kb,
        and the entire population (1000 diploids in each of four populations) were used.
        \label{fig:introf4}
    }
\end{figure}
